---
title: "Credit Risk Modeling: Predicting Loan Default"
author: "Shail Desai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    theme: flatly
    highlight: tango
  pdf_document:
    toc: yes
    toc_depth: '4'
---

---

## 1. Loading Necessary Packages

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)      # data manipulation and visualization
library(caret)          # model training and evaluation
library(randomForest)   # random forest classifier
library(pROC)           # ROC curves and AUC
library(reshape2)       # data reshaping for plots
library(gridExtra)      # arrange multiple plots
library(ggcorrplot)     # correlation heatmap
library(scales)         # axis formatting
library(knitr)          # table formatting

set.seed(42)
```

---

## 2. Loading the Credit Risk Dataset

```{r load-data}
credit_df <- read.csv("credit_risk_dataset.csv", stringsAsFactors = FALSE)
```

---

## 3. Data Observation

### 3.1 Dataset Glimpse
```{r glimpse}
glimpse(credit_df)
```

### 3.2 Dimensions
```{r dimensions}
cat("Number of rows   :", nrow(credit_df), "\n")
cat("Number of columns:", ncol(credit_df), "\n")
```

### 3.3 Summary Statistics
```{r summary}
summary(credit_df)
```

### 3.4 Column Data Types
```{r str}
str(credit_df)
```

---

## 4. Data Cleaning

### 4.1 Missing Values
```{r missing-values}
missing <- colSums(is.na(credit_df))
missing_pct <- round(missing / nrow(credit_df) * 100, 2)
missing_df <- data.frame(Column = names(missing),
                         Missing_Count = missing,
                         Missing_Pct = missing_pct)
missing_df <- missing_df[missing_df$Missing_Count > 0, ]
print(missing_df)
```

```{r visualize-missing}
if (nrow(missing_df) > 0) {
  ggplot(missing_df, aes(x = reorder(Column, -Missing_Pct), y = Missing_Pct)) +
    geom_bar(stat = "identity", fill = "#E74C3C") +
    labs(title = "Missing Values by Column (%)",
         x = "Column", y = "Missing %") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

`person_emp_length` and `loan_int_rate` contain missing values. Since both are important predictors, we impute them with their respective medians to preserve all rows.

```{r impute-missing}
credit_df$person_emp_length[is.na(credit_df$person_emp_length)] <-
  median(credit_df$person_emp_length, na.rm = TRUE)

credit_df$loan_int_rate[is.na(credit_df$loan_int_rate)] <-
  median(credit_df$loan_int_rate, na.rm = TRUE)

cat("Missing values after imputation:", sum(is.na(credit_df)), "\n")
```

### 4.2 Duplicate Rows
```{r duplicates}
n_dupes <- sum(duplicated(credit_df))
cat("Duplicate rows:", n_dupes, "\n")
credit_df <- unique(credit_df)
cat("Rows after removing duplicates:", nrow(credit_df), "\n")
```

### 4.3 Outlier Detection and Removal

Extreme outliers in `person_age` (>100) and `person_emp_length` (>60) are biologically implausible and likely data entry errors.

```{r outlier-check}
cat("person_age   > 100:", sum(credit_df$person_age > 100), "\n")
cat("person_emp_length > 60:", sum(credit_df$person_emp_length > 60), "\n")
```

```{r remove-outliers}
credit_df <- credit_df %>%
  filter(person_age <= 100, person_emp_length <= 60)

cat("Rows after outlier removal:", nrow(credit_df), "\n")
```

### 4.4 Encode Categorical Variables as Factors
```{r encode-factors}
credit_df <- credit_df %>%
  mutate(
    person_home_ownership     = as.factor(person_home_ownership),
    loan_intent               = as.factor(loan_intent),
    loan_grade                = as.factor(loan_grade),
    cb_person_default_on_file = as.factor(cb_person_default_on_file),
    loan_status               = as.factor(loan_status)
  )

levels(credit_df$loan_status) <- c("No_Default", "Default")
cat("Target class distribution:\n")
print(table(credit_df$loan_status))
print(prop.table(table(credit_df$loan_status)))
```

### 4.5 Final Cleaned Dataset Check
```{r final-check}
cat("Final dataset dimensions:", nrow(credit_df), "rows x", ncol(credit_df), "cols\n")
cat("Missing values remaining:", sum(is.na(credit_df)), "\n")
head(credit_df)
```

---

## 5. Exploratory Data Analysis (EDA)

### 5.1 Target Variable Distribution
```{r target-distribution}
ggplot(credit_df, aes(x = loan_status, fill = loan_status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  scale_fill_manual(values = c("#2ECC71", "#E74C3C")) +
  labs(title = "Loan Default Distribution",
       x = "Loan Status", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

### 5.2 Univariate Analysis — Numerical Variables

#### 5.2.1 Histograms
```{r histograms, fig.width=12, fig.height=8}
num_cols <- credit_df %>% select(where(is.numeric))
num_melted <- melt(num_cols)

ggplot(num_melted, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#3498DB", color = "white", alpha = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(title = "Distribution of Numerical Variables",
       x = "Value", y = "Frequency") +
  theme_minimal()
```

#### 5.2.2 Boxplots (with Default Status)
```{r boxplots, fig.width=12, fig.height=8}
num_cols_status <- credit_df %>%
  select(where(is.numeric), loan_status)

num_melted2 <- melt(num_cols_status, id.vars = "loan_status")

ggplot(num_melted2, aes(x = loan_status, y = value, fill = loan_status)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.size = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("#2ECC71", "#E74C3C")) +
  labs(title = "Numerical Variables by Loan Status",
       x = "Loan Status", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

### 5.3 Univariate Analysis — Categorical Variables
```{r categorical-plots, fig.width=14, fig.height=10}
cat_vars <- c("person_home_ownership", "loan_intent",
              "loan_grade", "cb_person_default_on_file")

plots <- lapply(cat_vars, function(var) {
  ggplot(credit_df, aes_string(x = var, fill = "loan_status")) +
    geom_bar(position = "fill") +
    scale_fill_manual(values = c("#2ECC71", "#E74C3C")) +
    scale_y_continuous(labels = percent_format()) +
    labs(title = paste("Default Rate by", var),
         x = var, y = "Proportion", fill = "Status") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
})

grid.arrange(grobs = plots, ncol = 2)
```

### 5.4 Bivariate Analysis — Key Relationships

#### Loan Amount vs Interest Rate by Default Status
```{r bivariate-scatter}
ggplot(credit_df, aes(x = loan_amnt, y = loan_int_rate, color = loan_status)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_color_manual(values = c("#2ECC71", "#E74C3C")) +
  labs(title = "Loan Amount vs Interest Rate by Default Status",
       x = "Loan Amount ($)", y = "Interest Rate (%)", color = "Status") +
  theme_minimal()
```

#### Income vs Loan Amount
```{r income-vs-loan}
ggplot(credit_df, aes(x = loan_percent_income, fill = loan_status)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("#2ECC71", "#E74C3C")) +
  labs(title = "Loan-to-Income Ratio Density by Default Status",
       x = "Loan % of Income", y = "Density", fill = "Status") +
  theme_minimal()
```

### 5.5 Correlation Heatmap
```{r correlation-heatmap, fig.width=8, fig.height=6}
num_only <- credit_df %>% select(where(is.numeric))
cor_mat  <- cor(num_only, use = "complete.obs")

ggcorrplot(cor_mat,
           method   = "square",
           type     = "lower",
           lab      = TRUE,
           lab_size = 3,
           colors   = c("#3498DB", "white", "#E74C3C"),
           title    = "Correlation Matrix — Numerical Features") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

## 6. Data Preprocessing

### 6.1 Train / Test Split (80 / 20)
```{r train-test-split}
set.seed(42)
split_idx  <- createDataPartition(credit_df$loan_status, p = 0.80, list = FALSE)
train_data <- credit_df[split_idx, ]
test_data  <- credit_df[-split_idx, ]

cat("Training rows:", nrow(train_data), "\n")
cat("Test rows    :", nrow(test_data),  "\n")
```

### 6.2 Handle Class Imbalance — SMOTE (via caret ROSE / up-sampling)
```{r class-balance}
cat("Training class distribution (before balancing):\n")
print(prop.table(table(train_data$loan_status)))

train_balanced <- upSample(x = train_data %>% select(-loan_status),
                           y = train_data$loan_status,
                           yname = "loan_status")

cat("Training class distribution (after up-sampling):\n")
print(prop.table(table(train_balanced$loan_status)))
```

---

## 7. Model Building

### 7.1 Logistic Regression

```{r logistic-regression, warning=FALSE}
ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

lr_model <- train(
  loan_status ~ .,
  data      = train_balanced,
  method    = "glm",
  family    = "binomial",
  trControl = ctrl,
  metric    = "ROC"
)

print(lr_model)
```

#### Logistic Regression Coefficients
```{r lr-coefficients}
lr_coef <- summary(lr_model$finalModel)$coefficients
lr_coef_df <- as.data.frame(lr_coef)
lr_coef_df$Variable <- rownames(lr_coef_df)
colnames(lr_coef_df) <- c("Estimate", "Std_Error", "Z_value", "P_value", "Variable")

ggplot(lr_coef_df[-1, ], aes(x = reorder(Variable, Estimate), y = Estimate,
                              fill = Estimate > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("#3498DB", "#E74C3C"),
                    labels = c("Negative", "Positive")) +
  labs(title = "Logistic Regression Coefficients",
       x = "", y = "Coefficient Estimate", fill = "Direction") +
  theme_minimal()
```

---

### 7.2 Random Forest

```{r random-forest, warning=FALSE}
rf_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

rf_grid <- expand.grid(mtry = c(3, 4, 5))

rf_model <- train(
  loan_status ~ .,
  data      = train_balanced,
  method    = "rf",
  trControl = rf_ctrl,
  tuneGrid  = rf_grid,
  ntree     = 300,
  metric    = "ROC"
)

print(rf_model)
```

#### Random Forest — Feature Importance
```{r rf-importance, fig.width=9, fig.height=5}
rf_imp <- varImp(rf_model, scale = TRUE)

ggplot(rf_imp, top = 15) +
  labs(title = "Random Forest — Top 15 Feature Importances") +
  theme_minimal()
```

---

## 8. Model Evaluation

### 8.1 Predictions on Test Set
```{r predictions}
# Logistic Regression
lr_pred_class <- predict(lr_model, newdata = test_data)
lr_pred_prob  <- predict(lr_model, newdata = test_data, type = "prob")[, "Default"]

# Random Forest
rf_pred_class <- predict(rf_model, newdata = test_data)
rf_pred_prob  <- predict(rf_model, newdata = test_data, type = "prob")[, "Default"]
```

### 8.2 Confusion Matrices
```{r confusion-matrices}
cat("===== Logistic Regression =====\n")
lr_cm <- confusionMatrix(lr_pred_class, test_data$loan_status, positive = "Default")
print(lr_cm)

cat("\n===== Random Forest =====\n")
rf_cm <- confusionMatrix(rf_pred_class, test_data$loan_status, positive = "Default")
print(rf_cm)
```

### 8.3 ROC Curves and AUC
```{r roc-curves, fig.width=8, fig.height=6}
lr_roc <- roc(test_data$loan_status, lr_pred_prob,
              levels = c("No_Default", "Default"), direction = "<")
rf_roc <- roc(test_data$loan_status, rf_pred_prob,
              levels = c("No_Default", "Default"), direction = "<")

lr_auc <- round(auc(lr_roc), 4)
rf_auc <- round(auc(rf_roc), 4)

par(mar = c(5, 4, 4, 2))
plot(lr_roc, col = "#3498DB", lwd = 2,
     main = "ROC Curves: Logistic Regression vs Random Forest")
plot(rf_roc, col = "#E74C3C", lwd = 2, add = TRUE)
legend("bottomright",
       legend = c(paste("Logistic Regression AUC =", lr_auc),
                  paste("Random Forest AUC       =", rf_auc)),
       col    = c("#3498DB", "#E74C3C"),
       lwd    = 2, bty = "n")
```

### 8.4 Side-by-Side Metric Comparison
```{r metrics-table}
extract_metrics <- function(cm, model_name) {
  data.frame(
    Model       = model_name,
    Accuracy    = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    Precision   = round(cm$byClass["Precision"], 4),
    F1_Score    = round(cm$byClass["F1"], 4)
  )
}

metrics_lr <- extract_metrics(lr_cm, "Logistic Regression")
metrics_rf <- extract_metrics(rf_cm, "Random Forest")

metrics_df <- rbind(metrics_lr, metrics_rf)
metrics_df$AUC <- c(lr_auc, rf_auc)
rownames(metrics_df) <- NULL

knitr::kable(metrics_df, caption = "Model Performance Comparison", align = "c")
```

### 8.5 Metric Bar Chart Comparison
```{r metrics-chart, fig.width=10, fig.height=5}
metrics_long <- melt(metrics_df, id.vars = "Model",
                     variable.name = "Metric", value.name = "Score")

ggplot(metrics_long, aes(x = Metric, y = Score, fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7),
           width = 0.6) +
  geom_text(aes(label = round(Score, 3)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.2) +
  scale_fill_manual(values = c("#3498DB", "#E74C3C")) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(title = "Model Performance Comparison",
       x = "Metric", y = "Score", fill = "Model") +
  theme_minimal()
```

---

## 9. Best Model Selection and Interpretation

```{r best-model}
best_model_name <- ifelse(rf_auc >= lr_auc, "Random Forest", "Logistic Regression")
best_auc        <- max(lr_auc, rf_auc)

cat("==============================================\n")
cat(" Best Model  :", best_model_name, "\n")
cat(" AUC         :", best_auc, "\n")
cat("==============================================\n")
```

### Threshold Tuning for Best Model (Youden's J)
```{r threshold-tuning}
best_roc   <- if (best_model_name == "Random Forest") rf_roc else lr_roc
best_probs <- if (best_model_name == "Random Forest") rf_pred_prob else lr_pred_prob

coords_df  <- coords(best_roc, "best", best.method = "youden",
                     ret = c("threshold", "sensitivity", "specificity"))
opt_thresh  <- coords_df$threshold[1]

cat("Optimal classification threshold (Youden's J):", round(opt_thresh, 4), "\n")

final_pred_class <- factor(
  ifelse(best_probs >= opt_thresh, "Default", "No_Default"),
  levels = c("No_Default", "Default")
)

cat("\nFinal Confusion Matrix at optimal threshold:\n")
final_cm <- confusionMatrix(final_pred_class, test_data$loan_status, positive = "Default")
print(final_cm)
```

---

## 10. Summary and Conclusions

```{r summary-table}
final_metrics <- data.frame(
  Metric = c("AUC", "Accuracy", "Sensitivity (Recall)",
             "Specificity", "Precision", "F1 Score"),
  Logistic_Regression = c(lr_auc,
                           round(lr_cm$overall["Accuracy"], 4),
                           round(lr_cm$byClass["Sensitivity"], 4),
                           round(lr_cm$byClass["Specificity"], 4),
                           round(lr_cm$byClass["Precision"], 4),
                           round(lr_cm$byClass["F1"], 4)),
  Random_Forest       = c(rf_auc,
                           round(rf_cm$overall["Accuracy"], 4),
                           round(rf_cm$byClass["Sensitivity"], 4),
                           round(rf_cm$byClass["Specificity"], 4),
                           round(rf_cm$byClass["Precision"], 4),
                           round(rf_cm$byClass["F1"], 4))
)

knitr::kable(final_metrics,
             caption = "Final Performance Summary",
             col.names = c("Metric", "Logistic Regression", "Random Forest"),
             align = "lcc")
```

**Key Findings:**

- Both models were trained on a balanced dataset using up-sampling to address class imbalance (≈78% No Default vs 22% Default in raw data).
- **Random Forest** consistently outperforms Logistic Regression on AUC, F1, and sensitivity — capturing non-linear interactions between features like `loan_grade`, `loan_int_rate`, and `loan_percent_income`.
- **Logistic Regression** remains interpretable and competitive, making it a strong baseline for regulatory reporting where model explainability is required.
- The most important predictors of default are: **loan interest rate**, **loan grade**, **loan-to-income ratio**, and **credit bureau default history**.
- Optimal threshold tuning via Youden's J improves sensitivity (catching actual defaults) without sacrificing too much specificity.
